%%%%%%%%%%%%%%%%%%%%% chapter15.tex %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Chapter 15: 70B Deployment
%
%%%%%%%%%%%%%%%%%%%%%%%% Springer-Verlag %%%%%%%%%%%%%%%%%%%%%%%%%%

\chapter{70B Model Deployment}
\label{ch:70b-deployment}

\abstract*{This chapter synthesizes everything from Part III into a complete 70B deployment. We cover the full setup with multi-GPU inference, tenant management, billing integration, and production monitoring. By the end, you'll have Control Plane v0.3---a multi-tenant AI platform serving enterprise customers.}

\abstract{This chapter synthesizes everything from Part III into a complete 70B deployment. We cover the full setup with multi-GPU inference, tenant management, billing integration, and production monitoring. By the end, you'll have Control Plane v0.3---a multi-tenant AI platform serving enterprise customers.}

% =============================================================================
\section{Deployment Architecture}
\label{sec:70b-architecture}
% =============================================================================

% TODO: Full architecture diagram
% - Multiple inference nodes
% - Control plane cluster
% - Database
% - Monitoring

% =============================================================================
\section{Infrastructure Setup}
\label{sec:infra-setup}
% =============================================================================

% TODO: Setting up the infrastructure
% - GPU nodes
% - Kubernetes or Docker Swarm
% - Networking

% =============================================================================
\section{Control Plane v0.3 Configuration}
\label{sec:cpv03-config}
% =============================================================================

% TODO: Complete configuration
% - All features enabled
% - Production tuning

% =============================================================================
\section{Deployment Checklist}
\label{sec:deployment-checklist}
% =============================================================================

% TODO: Pre-launch checklist
% - Security
% - Monitoring
% - Backup
% - Documentation

% =============================================================================
\section{Operations Runbook}
\label{sec:runbook}
% =============================================================================

% TODO: Day-to-day operations
% - Scaling
% - Updates
% - Incident response

% =============================================================================
\section{Summary}
\label{sec:ch15-summary}
% =============================================================================

\begin{important}{Key Takeaways}
\begin{itemize}
\item 70B deployment requires careful planning
\item Multi-GPU coordination is handled by vLLM
\item Control Plane v0.3 provides full multi-tenancy
\item Operational procedures are essential
\end{itemize}
\end{important}

\begin{svgraybox}
\textbf{Part III Complete!}

You now have Control Plane v0.3, a multi-tenant AI platform with:
\begin{itemize}
\item Tenant isolation and management
\item Usage tracking and billing
\item Multi-GPU distributed inference
\item Intelligent model routing
\item Production deployment patterns
\end{itemize}

In Part IV, we'll push to 400B models and build CodeLab, a commercial coding assistant.
\end{svgraybox}

\section*{Problems}
\addcontentsline{toc}{section}{Problems}

\begin{prob}
\label{prob:ch15-deployment}
Deploy a complete 70B inference stack with at least 3 tenants. Document the process and measure end-to-end latency.
\end{prob}

\input{chapters/chapter16/references}
