%%%%%%%%%%%%%%%%%%%%% chapter14.tex %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Chapter 14: Model Routing and Selection
%
%%%%%%%%%%%%%%%%%%%%%%%% Springer-Verlag %%%%%%%%%%%%%%%%%%%%%%%%%%

\chapter{Model Routing and Selection}
\label{ch:routing}

\abstract*{Not every request needs your largest model. This chapter covers intelligent model routing based on request complexity, cost optimization through model selection, A/B testing different models, and building a routing engine that balances quality and cost. Smart routing can reduce costs by 50\% while maintaining quality.}

\abstract{Not every request needs your largest model. This chapter covers intelligent model routing based on request complexity, cost optimization through model selection, A/B testing different models, and building a routing engine that balances quality and cost. Smart routing can reduce costs by 50\% while maintaining quality.}

% =============================================================================
\section{Why Route Between Models?}
\label{sec:why-routing}
% =============================================================================

% TODO: Motivation
% - Cost optimization
% - Latency requirements
% - Quality matching

% =============================================================================
\section{Routing Strategies}
\label{sec:routing-strategies}
% =============================================================================

\subsection{Rule-Based Routing}
\label{subsec:rule-routing}

% TODO: Simple rules
% - Task type -> model
% - Token length -> model

\subsection{Complexity-Based Routing}
\label{subsec:complexity-routing}

% TODO: Estimating complexity
% - Prompt analysis
% - User tier

\subsection{Cost-Aware Routing}
\label{subsec:cost-routing}

% TODO: Cheapest model that works

% =============================================================================
\section{Router Implementation}
\label{sec:router-impl}
% =============================================================================

\begin{programcode}{Model Router}
\begin{lstlisting}[language=Go]
// internal/router/router.go

type ModelRouter interface {
    SelectModel(ctx context.Context, req RoutingRequest) (string, error)
}

type RoutingRequest struct {
    Prompt     string
    MaxTokens  int
    UserTier   string
    TaskType   string
    LatencySLA time.Duration
}

type RuleBasedRouter struct {
    rules []RoutingRule
}

type RoutingRule struct {
    Condition func(RoutingRequest) bool
    Model     string
    Priority  int
}
\end{lstlisting}
\end{programcode}

% =============================================================================
\section{A/B Testing Models}
\label{sec:ab-testing}
% =============================================================================

% TODO: Testing model performance
% - Traffic splitting
% - Metrics collection
% - Statistical analysis

% =============================================================================
\section{Fallback Strategies}
\label{sec:fallback}
% =============================================================================

% TODO: When primary model fails
% - Automatic failover
% - Quality degradation

% =============================================================================
\section{Summary}
\label{sec:ch14-summary}
% =============================================================================

\begin{important}{Key Takeaways}
\begin{itemize}
\item Smart routing can reduce costs by 30-50\%
\item Match model capability to request complexity
\item A/B testing validates routing decisions
\item Always have fallback models
\end{itemize}
\end{important}

\section*{Problems}
\addcontentsline{toc}{section}{Problems}

\begin{prob}
\label{prob:ch14-router}
Implement a model router that selects between 7B, 30B, and 70B models based on prompt length, user tier, and task type.
\end{prob}

\input{chapters/references14}
