%%%%%%%%%%%%%%%%%%%%% chapter18.tex %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Chapter 18: Production Operations and Control Plane v1.0
%
%%%%%%%%%%%%%%%%%%%%%%%% Springer-Verlag %%%%%%%%%%%%%%%%%%%%%%%%%%

\chapter{Production Operations and Control Plane v1.0}
\label{ch:operations}

\abstract*{The final chapter covers production operations at scale: advanced monitoring, automated incident response, capacity planning, and the complete Control Plane v1.0. We discuss reliability engineering practices for AI infrastructure and provide a comprehensive operational playbook.}

\abstract{The final chapter covers production operations at scale: advanced monitoring, automated incident response, capacity planning, and the complete Control Plane v1.0. We discuss reliability engineering practices for AI infrastructure and provide a comprehensive operational playbook.}

% =============================================================================
\section{Control Plane v1.0 Overview}
\label{sec:cpv10-overview}
% =============================================================================

% TODO: Complete feature set
% - All features from v0.1-v0.3
% - Production hardening
% - Enterprise features

\begin{svgraybox}
\textbf{Control Plane v1.0 Features:}

\textbf{From v0.1 (Foundations):}
\begin{itemize}
\item Backend abstraction
\item Health checks
\item Prometheus metrics
\end{itemize}

\textbf{From v0.2 (Production):}
\begin{itemize}
\item Authentication
\item Rate limiting
\item Caching
\item Request queue
\end{itemize}

\textbf{From v0.3 (Multi-Tenant):}
\begin{itemize}
\item Tenant isolation
\item Billing integration
\item Distributed inference
\item Model routing
\end{itemize}

\textbf{New in v1.0:}
\begin{itemize}
\item Distributed tracing
\item Automated failover
\item Capacity planning
\item Cost optimization
\end{itemize}
\end{svgraybox}

% =============================================================================
\section{Advanced Observability}
\label{sec:advanced-observability}
% =============================================================================

% TODO: Beyond basic metrics
% - Distributed tracing
% - Log aggregation
% - Anomaly detection

\subsection{Distributed Tracing}
\label{subsec:tracing}

% TODO: Request tracing
% - OpenTelemetry
% - Jaeger/Zipkin

\subsection{SLOs and SLIs}
\label{subsec:slo-sli}

% TODO: Service level objectives
% - Defining SLOs
% - Measuring SLIs
% - Error budgets

% =============================================================================
\section{Incident Management}
\label{sec:incident-management}
% =============================================================================

% TODO: When things go wrong
% - Detection
% - Response
% - Recovery
% - Post-mortem

% =============================================================================
\section{Capacity Planning}
\label{sec:capacity-planning}
% =============================================================================

% TODO: Planning for growth
% - Traffic forecasting
% - Hardware procurement
% - Scaling triggers

% =============================================================================
\section{Automated Operations}
\label{sec:automation}
% =============================================================================

% TODO: Reducing toil
% - Auto-scaling
% - Auto-recovery
% - Deployment automation

% =============================================================================
\section{Cost Optimization}
\label{sec:cost-optimization}
% =============================================================================

% TODO: Managing costs at scale
% - Utilization optimization
% - Spot instances
% - Model efficiency

% =============================================================================
\section{Security Hardening}
\label{sec:security-hardening}
% =============================================================================

% TODO: Production security
% - Network security
% - Secrets management
% - Audit logging

% =============================================================================
\section{Operational Playbook}
\label{sec:playbook}
% =============================================================================

% TODO: Runbook compilation
% - Deployment
% - Scaling
% - Incidents
% - Maintenance

% =============================================================================
\section{Conclusion}
\label{sec:conclusion}
% =============================================================================

% TODO: Book conclusion
% - What you've built
% - Where to go next
% - The future of self-hosted AI

\begin{svgraybox}
\textbf{Congratulations!}

You've built a complete AI inference platform, from a single 7B model on a laptop to a commercial 400B inference lab. Along the way, you've learned:

\begin{itemize}
\item How inference works at every scale
\item Hardware selection and optimization
\item Production infrastructure patterns
\item Multi-tenant platform design
\item Commercial product development
\end{itemize}

The control plane you've built is production-ready. The knowledge you've gained puts you at the forefront of AI infrastructure engineering.

The future of AI is self-hosted. Go build it.
\end{svgraybox}

% =============================================================================
\section{Summary}
\label{sec:ch18-summary}
% =============================================================================

\begin{important}{Key Takeaways}
\begin{itemize}
\item Control Plane v1.0 is production-complete
\item Observability goes beyond basic metrics
\item Automation reduces operational burden
\item Security and reliability are non-negotiable
\item You now have the skills to run world-class AI infrastructure
\end{itemize}
\end{important}

\section*{Problems}
\addcontentsline{toc}{section}{Problems}

\begin{prob}
\label{prob:ch18-slo}
Define SLOs for your inference platform including availability, latency, and error rate. Implement dashboards to track them.
\end{prob}

\begin{prob}
\label{prob:ch18-runbook}
Create a complete operational runbook for your deployment including procedures for deployment, scaling, and incident response.
\end{prob}

\input{chapters/chapter19/references}
